---
title: "Analyzing English Premier League Match Data"
subtitle: "Seasons 2021-23"
author: 
  - name: "Mildcats"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project proposal outlining the objective, datasets, research questions and plan"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: r-setup-load-pkgs
#| message: false

if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(here,
               tidyverse,
               dplyr,
               readr)
```

## Introduction

The English Premier League is consistently ranked as the most viewed sports league in the world. With widespread popularity and fanfare, the Premier League is an interesting subject to further explore using data methods, to uncover interesting patterns and phenomenon that fans may otherwise miss.

Many fans also bet heavily on the outcome of matches, with the Premier League being one of the most attractive events for bettors. Thus, analyzing match data in sync with betting odds data, we may also find some interesting results that may be of relevance to fans who are risking money.

After conducting data analysis, we aim to visualize and present it in an intriguing, yet easy to understand fashion, in order to make it accessible to a wide range of audience.

## Datasets

### Match Data

The core data-set for our project is the the [Premier League Match Data](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-04-04/readme.md) for the 2021-22 season. This data-set will be [extended](https://www.kaggle.com/datasets/evangower/premier-league-2022-2023) to cover matches from years 2022-2023.

As sports fans, we often find ourselves absorbed by the adrenaline and pace of the game, but taking a broader, objective approach can help us answer "difficult" questions like -

-   Why did my team lose this game (that we were sure we would win)

-   Is there such a thing as home ground advantage? Do some teams objectively and quantifiably perform better on home grounds, or are there teams which are agnostic to this?

Hence, this data-set stood out in it's ability to be an interesting case-study towards building an engaging project, making it our choice for this project.

#### Data Description

| **Column Name** | **Description**                                   | **Type**    | **Usage in Project**                                |
|-----------------|---------------------------------------------------|-------------|-----------------------------------------------------|
| **Date**        | Date of the match                                 | Datetime    | Used to align matches with weather and betting data |
| **HomeTeam**    | Name of the home team                             | Categorical | Analyze home team performance                       |
| **AwayTeam**    | Name of the away team                             | Categorical | Compare with home team performance                  |
| **FTHG**        | Full-time home goals                              | Integer     | Key for match outcome analysis                      |
| **FTAG**        | Full-time away goals                              | Integer     | Key for match outcome analysis                      |
| **FTR**         | Full-time result (H=Home Win, D=Draw, A=Away Win) | Categorical | Evaluate outcome trends and betting accuracy        |
| **HS**          | Home team shots                                   | Integer     | Analyze team performance metrics                    |
| **AS**          | Away team shots                                   | Integer     | Analyze team performance metrics                    |

We currently have the unprocessed, uncombined data-set as *soccer22-23.csv* in our raw data directory, but this will be merged to form a *soccer21-23.csv* dataset (a total of **760** rows).

```{r}
#| label: load-match-data
#| message: false

match_data <- read.csv(here("data", "raw_data", "soccer21-22_with_location.csv"))
glimpse(match_data)
```

### Augmenting Datasets

Match data is planned to be supplemented with publicly sourced weather and betting odds data to form the cumulative, unprocessed raw data.

#### Weather Data

This data was chosen as a potential augmentation to the match data as weather events significantly affect the playing conditions. This can influence results of the game, often in an unexpected way.

Some teams might have strategies, or players that do well in wet conditions that slow down the overall pace of the game, and having this data in our analysis would help us test various hypothesis that can form the crux of our research questions.

##### Data Description

| **Column Name** | **Description**                  | **Type** | **Usage in Project**                                         |
|-----------------|----------------------------------|----------|--------------------------------------------------------------|
| **datetime**    | Timestamp of weather data        | Datetime | Align weather with match date                                |
| **tempmax**     | Maximum temperature on match day | Float    | Analyze impact of temperature on match outcomes              |
| **tempmin**     | Minimum temperature on match day | Float    | Analyze impact of temperature fluctuations                   |
| **precip**      | Precipitation (in mm)            | Float    | Evaluate how weather impacts match performance               |
| **windspeed**   | Speed of wind on match day       | Float    | Analyze influence of wind on match results                   |
| **humidity**    | Humidity on match day            | Float    | Investigate correlations between humidity and match outcomes |
| **visibility**  | Visibility during match          | Float    | Explore effects of low visibility on performance             |

**Note** - The city-wise data will be combined/alternatively sourced to form a master weather data-set.

```{r}
#| label: load-dataset-weather
#| message: false

weather_leeds <- read.csv(here("data", "raw_data", "weather21-22", "Leeds.csv"))
glimpse(weather_leeds)
```

#### Betting Odds Data

This data-set was included as part of our analysis to answer a few key questions -

-   Is there a way to link pre-published betting odds with the outcome of league games?

-   How often are these odds pointing in the wrong direction (e.g. Underdog wins, favorite loses)

-   Is there a way to quantify and link a team's odds progression through the season to their performance

Finally, we wanted to check if it is the "house" (read bookmaker) always wins? Or is there a way for a sharp (seasoned sports bettor) to outsmart the book.

##### Data Descripton

| **Column Name** | **Description**                | **Type**    | **Usage in Project**                                            |
|-----------------|--------------------------------|-------------|-----------------------------------------------------------------|
| **HomeTeam**    | Name of the home team          | Categorical | Compare betting odds to actual results                          |
| **AwayTeam**    | Name of the away team          | Categorical | Compare betting odds to actual results                          |
| **B365H**       | Bet365 odds for home team win  | Float       | Analyze accuracy of betting odds                                |
| **B365D**       | Bet365 odds for a draw         | Float       | Analyze odds dynamics for tied matches                          |
| **B365A**       | Bet365 odds for away team win  | Float       | Compare odds with actual match outcome                          |
| **MaxH**        | Maximum odds for home team win | Float       | Track the highest betting odds and its correlation with results |
| **MaxA**        | Maximum odds for away team win | Float       | Track the highest betting odds and its correlation with results |

**Note** - The betting odds data from the 2 seasons (21-22, 22-23) will be combined to form a master betting odds data-set (760 rows)

```{r}
#| label: load-dataset-betodds
#| message: false

bet_odds <- read.csv(here("data", "raw_data", "betting_odds21-22.csv"))
glimpse(bet_odds)
```

#### Data Source and Ethical Considerations

-   **Match Data**

    -   **Source** -

        -   This data was sourced from [Evan Gower](https://www.kaggle.com/evangower) via [Kaggle](https://www.kaggle.com/datasets/evangower/premier-league-match-data) (originally referenced from the [tidytuesday list of datasets](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023#readme))

        -   The [extension](https://www.kaggle.com/datasets/evangower/premier-league-2022-2023) (to the 22-23 season) was also sourced from the same user

    -   **Collection Method -**

        -   The provenance of the cleaned match data is from the [FootballData UK](https://www.football-data.co.uk/englandm.php) website.

        -   We could not identify explicitly how this organization collects data, but it seems that the incoming data stream relies on live match data from gambling websites/book-makers.

    -   **Ethical Considerations -**

        -   The data on Kaggle is offered openly, but the copyright lies with the original authors (<https://www.football-data.co.uk/>)

        -   Since the original author provides the data publicly, we are within our rights to use this data for a non-commercial use case i.e. our project.

-   **Weather**

    -   **Source** - Data for the weather during matches has been retrieved from the weather data vendor [Visual Crossing](https://www.visualcrossing.com/weather-data).

    -   **Collection Method -** This provider compiles data by collecting raw measurements from observation stations located all across the world. Given that the locations of interest for us are all major cities in the U.K., we can be quite confident in the quality of this data.

    -   **Ethical Considerations -**

        -   The weather data was collected through legal and legitimate means by using the Visual Crossing platform, by adhering to the platform's terms and conditions.

        -   Since there was a limit of 1000 record retrieval cap in a 24 hour period, the workload was divided among 6 team members.

        -   The data was not manipulated at any point, thus allowing transperency and clear documentation of the process.

        -   The data collected will solely be used for the intended purpose of this project.

-   **Betting Odds**

    -   **Source** -

        -   This data was sourced from [FootballData UK](https://www.football-data.co.uk/englandm.php)

        -   Co-incidentally, this is also the primary data source for our cleaned match data

    -   **Collection Method -**

        -   We could not identify explicitly how this organization collects data, but it seems that the incoming data stream relies on live match data from gambling websites/book-makers.

    -   **Ethical Considerations -**

        -   Since the original author provides the data publicly, we are within our rights to use this data for a non-commercial use case i.e. our project. No ethical conflicts or ambiguity here.

## Research Questions

-   **How do extreme pre-match weather conditions (such as heavy rain, strong winds, or cold temperatures) influence the likelihood of surprising outcomes, such as underdog victories, and how do betting odds before the match reflect these potential surprises compared to historical odds for the same teams?**

    -   **Rationale**:

        -   We aim to explore if extreme weather conditions lead to more unexpected outcomes, such as the underdogs (as per the pre match betting line) winning.

        -   It also examines how pre-match betting odds account for these potential disruptions, by comparing them to odds for the same teams in previous matches under normal conditions.

    -   **Datasets involved**:

        -   Match data

        -   Weather data

        -   Pre-match betting odds.

-   **Is there a quantifiable home team advantage? And is this reflected in the pre-match betting odds?**

    -   **Rationale -**

        -   We aim to trace whether there is a significant uptick in a team's perceived performance when they are playing on their home ground

        -   For similarly ranked teams, with similar records, do their betting odds reflect a perceived advantage when playing at home vs playing away?

        -   If the odds are indicative of this advantage, how often does this translate to the actual result

    -   **Datasets involved**:

        -   Match data

        -   Pre-match betting odds.

-   **How accurate are the pre match betting odds?**

    -   **Rationale -**

        -   We aim to perform a simple accuracy analysis of pre match betting odds on the match outcome

        -   We aim to track the sensitivity of a team's rolling performance to the betting odds essentially tracking whether there is a recency bias built into betting odds

    -   **Datasets involved**:

        -   Match data

        -   Pre-match betting odds.

## Analysis Plan

-   Before we begin the analysis, the following pre-processing steps need to be performed -
    -   **Extending Match Data** - We need to merge the match data from 21-22 with 22-23 seasons to give us a parent match data-set
    -   **Flattening and Normalizing Weather Data -** Our weather data is currently city-wise, we need to aggregate this into a single data-set, and create a *foreign key* or identifier which will help us join it with the match data
        -   Additional statistical metrics might need to be calculated - Average Temperature, Precipitation etc.
    -   **Betting Odds Data** - Since none of us have experience with sports betting, we need to understand and isolate the important data points vis-a-vis bookmaking and betting odds and merge them into the match data-set.
-   The ETL has been completed via **etl\\dataPreprocessing.qmd** to extract the merged dataset that will be used as the main data source for this project - **soccer_main.csv**
